A Short and Mechanized Consistency Proof for Dependent Type Theory

### 2024-02-18 paper/paper.tex
comments throughout the paper

%%
%% This is file `sample-acmsmall.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `acmsmall')
%%
%% IMPORTANT NOTICE:
%%
%% For the copyright see the source file.
%%
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-acmsmall.tex.
%%
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%%
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart} or whichever template is required
%% for your publication.
%%
%%
\documentclass[acmsmall]{acmart}
\usepackage{ottalt}
\usepackage{minted}
\usepackage{xspace}
\usepackage{tcolorbox}
\usepackage[para]{footmisc}
\definecolor{lightgray}{gray}{0.75}
\newcommand{\dotv}[2]{\href{#1}{\texttt{#1}}{\texttt{:#2}}}
\newcommand{\lang}{$\lambda^H$\xspace}
\inputott{rules}
%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

@@

The consistency proofs of various dependently typed systems, including Martin-Lof's
type theory and the Calculus of Constructions, have long been
available in the literature. In particular, recent works such as \citet{nbeincoq},
\citet{decagda}, and \citet{martin-lof-a-la-coq} mechanize stronger
results related to the decidability of type conversion or type
checking of dependently typed systems, from which consistency can be
derived as a corollary.

The underlying technique of the forementioned
works is proof by logical relation,
which involves interpreting types as reducibility predicates,
representing sets of terms satisfying certain properties with respect
to the reduction relation. While the proof technique and the
consistency result for dependent types are both well-established,
there is a severe lack of rigorous and accessible material that shows
how proof by logical relation can be applied to dependently typed
systems.

Introductory materials about logical relations such as
\citet{skorstengaard2019introduction}, \citet{harper2016practical}
start from logical relations for simply typed languages, and
eventually extend the technique to build a relational model for System
F in order to derive
parametricity. \citet{harperkripke} gives a gentle introduction on how
logical relations for closed terms can be extended to talk about open
terms so we can derive properties such as normalization for well-typed
open terms. Overall, the introductory texts about logical relations
cover systems and properties with varying degrees of complexity, from
simply typed to polymorphicly typed, logical predicate to logical
equivalence, closed terms to open terms.

@@

The goal of this paper is to give a tutorial on proof by logical relation for
dependent types in a simple and digestable format.
The system we present is small but covers
important features that are only available or require special treatments in
dependently typed systems. We choose boolean
types over natural numbers as our base type for simplicity, but include an
intensional identity type in our type system to show how indexed
types are treated in a logical relational proof. Unlike
\citet{nbeincoq,decagda,martin-lof-a-la-coq} but similar to
\citet{anand2014towards}, we include an infinite hierarchy of
universes to not only support type-level computations, but also avoid
the unnecessary code duplication pointed out by \citet{nbeincoq} when
big and small types are treated non-uniformly.

@@

In Section~\ref{sec:extension}, we extend our logical
relation to prove the stronger property that every well-typed term
has a normal form. Most interestingly, we find that extending the
language with
$\eta$ rule for functions and the metatheoretic result to normalization for open terms are
mostly orthogonal to the fact that the language is dependently
typed. In our mechanized proof, the overall
structure of the logical relation remains unchanged, and the additional
proof obligations are mostly related to properties about the untyped
lambda calculus, all of which can be derived through syntactic
means.
In other words, once we have established the base technique for
proof by logical relation for dependent types, we can factor out the
complexities of an extension the are not specifically related to dependent
types; such an extension, if desired, may be studied in the context of
a simply typed language and later ported into a dependently typed setting.

@@

@@

\mathit{Contexts}\\
[[G]]       & ::= & [[empty]]\ |\ [[G ++ A]] &  \\ \\
\mathit{Terms}\\
[[a]],[[b]],[[c]],[[t]],[[p]],[[A]],[[B]] & ::= & [[Set i]]\ |\ [[var i]]\  |\ [[Void]]
                  & \mbox{universes, variables, empty type} \\
            & |   & [[Pi A B]]\ |\ [[\ A a]]\ |\ [[a b]]
                  & \mbox{function types, abstractions, applications} \\
            & |   & [[a ~ b : A ]]\ |\  [[refl]]\ |\ [[J t a b p]]
                  & \mbox{equality types, reflexivity proof, J eliminator} \\
            & |   & [[Bool]]\ |\  [[true]]\ |\  [[false]]\ % |\  [[if a b0 b1]]
                  & \mbox{boolean type, true, false} \\
            & |   & [[if a b0 b1]]
                  & \mbox{if} \\ \\
\mathit{Renaming}\\
[[xi]] & \in & [[SNat -> SNat]] & \\ \\
\mathit{Substitution}\\
[[rho]] & \in & [[SNat -> STm]] &
\end{array}
\]
  \caption{Syntax of \lang}
  \label{fig:syntax}
\end{figure}

@@

The syntax of \lang can be found in Figure~\ref{fig:syntax}. We use
the unscoped de Bruijn representation for both our Coq development and
the informal presentation in our paper. We find de Bruijn
representation advantageous for specification since it leaves very
little ambiguity about the variable freeness side
conditions, making our proof more easily reproducible. Furthermore, as we discuss in
Section~\ref{sec:automation}, de Bruijn representation is much more
amenable to automated reasoning in proof assistants.
We adopt from \citet{autosubst} the notations for simultaneous
renaming, substitution, and other auxiliary definitions related to our
syntax, summarized in Figure~\ref{fig:auxdef}. We omit most of the
definitions of renaming and substitution and only show the definition
of a few representative cases of substitution.
\begin{figure}[ht]
  \begin{equation*}
    \begin{split}
      \begin{array}{lll}
        \mathit{IdentityRen} \\
        [[id i]] & := & [[i]] \\ \\
        \mathit{IdentityTm} \\
        [[idtm i]] & := & [[var i]] \\ \\
        \mathit{ConsRen ([[xi .: n]])} \\
        ([[xi .: n]])([[0]]) & := & [[n]] \\
        ([[xi .: n]])([[Suc i]]) & := & [[xi i]] \\ \\
        \mathit{ConsSubst ([[rho .: a]])} \\
        ([[rho .: a]])([[0]]) & := & [[a]] \\
        ([[rho .: a]])([[Suc i]]) & := & [[rho i]] \\ \\
        \mathit{Curried\ Add} \\
        [[up n i]] & := & [[n]] + [[i]]
      \end{array}
    \end{split}
    \qquad \qquad
    \begin{split}
      \begin{array}{lll}
        \mathit{UpRen} \\
        [[( up xi ) 0]] & := & [[0]] \\
        [[( up xi ) Suc i]] & := & [[ Suc ren xi i ]] \\ \\
        \mathit{Renaming ([[a < xi >]])} \\
        [[(Pi A B) < xi >]] & := & [[Pi A < xi > B < up xi >]] \\
        [[(a b) < xi >]] & := & [[a < xi > (# b < xi > #)]] \\
        \ldots \\ \\
        \mathit{UpTm} \\
        [[( up rho ) 0]] & := & [[var 0]] \\
        [[( up rho ) Suc i]] & := & [[ rho i < up 1 > ]] \\ \\
        \mathit{Substitution ([[ a { rho }  ]])} \\
        [[var i { rho }  ]] & := &  [[rho i]] \\
        [[(Pi A B) { rho }]] & := & [[Pi A { rho } B { up rho }]] \\
        [[(a b) { rho }]] & := & [[a { rho } (# b { rho } #)]] \\
        \ldots
      \end{array}
    \end{split}
  \end{equation*}
  \caption{Auxiliary Functions over Syntax}
  \label{fig:auxdef}
\end{figure}

@@

% Figure~\ref{fig:auxdef} shows the auxiliary definitions over the term
% syntax, including renaming, substitution, and operations over
% the typing context or substitution.
As a
dependent type theory, terms and types are collapsed into the same
syntactic category. Dependent functions take the form $[[Pi A B]]$ and
we use the notation $[[A -> B]]$ when the output type $[[B]]$ is not
dependent on the input variable. $[[Set i]]$ represents the universe
type where $[[i]]$ ranges over the set of natural numbers.
 Finally,
\lang also includes an intensional identity type $[[a ~ b : A]]$ whose
proofs can be eliminated by the J-eliminator $[[J t a b p]]$, where
$[[p]]$ is an equality proof between $[[a]]$ and $[[b]]$, and $[[t]]$
is the term whose type is to be casted.

\lang is expressive enough to support large
elimination, the ability to compute a type using a term as input. For
example, the function $[[\ Bool if var 0 Bool Bool -> Bool]]$ returns
either $[[Bool]]$ or $[[Bool -> Bool]]$ depending on whether the input
is $[[true]]$ or $[[false]]$.

@@

Figure~\ref{fig:par} shows the definition of the parallel reduction
relation, which takes the form $[[a => b]]$. We use $[[a =>+ b]]$ to
represent its transitive and reflexive closure, which in turn allows us to define
the coherence relation $[[a <=> b]]$.
We say that two terms $[[a]]$
and $[[b]]$ are coherent if they can eventually reduce to some common
term $[[c]]$ through parallel reduction. The symmetric notation of
coherence suggests that it is an equivalence relation.

@@

From Lemma~\ref{lemma:coherencerefl}, \ref{lemma:coherencesym}, and
\ref{lemma:coherencetrans}, we conclude that coherence is indeed an
equivalence relation. It is then trivial to derive that coherence is
exactly the same as the untyped $\beta$-equivalence.

@@

Figure~\ref{fig:typing} gives the full typing rules for
\lang{}. The premises wrapped in \colorbox{lightgray}{gray} boxes can be shown to be
admissible syntactically, though some of them are required to
strengthen the inductive hypothesis of the fundamental theorem.
In \rref{T-Var}, $[[G i]]$ is the partial function defined
with the equations $[[ (G ++ A)  0 ]] := [[A]]$ and $[[ (G ++ A) Suc
i ]] := [[G i]] $. The precondition $[[i < | G | ]]$, where $[[ | G |
]]$ represents the length of the context $[[G]]$, ensures that $[[G
i]]$ has a defined output.
\Rref{T-Conv} uses the definition of coherence from earlier
as our equality judgment for type conversion. The use of an untyped
relation for type conversion makes our formulation slightly different
from languages such as MLTT~\citep{Martin-Lof-1973}, where the judgmental equality
takes the form $\Gamma \vdash a \equiv b : A$, which implies $[[G |- a :
A]]$ and $[[G |- b : A]]$, the well-typedness of $[[a]]$ and $[[b]]$.
\citet{siles2012pure} shows the equivalence of Barendregt's
Pure Type System~\citep{barendregt1991introduction}, which employs
untyped equality, and its variant that uses typed judgmental
equality. It is trivial to embed a system with judgmental
equality to a system with untyped equality by erasing typing
information. As a result, it is easy to port our consistency result from our
type system to a variant with typed judgmental equality
as long as the typed system does not include $\eta$ laws
that would require type annotations (e.g. the $\eta$-law for unit
types). % and we discuss in Section~\ref{sec:extension} how we can extend
% our proof to handle those rules.

@@

Finally, since our system has an infinite universe hierarchy, we can
present the system à la Russell by using the same judgment form $[[G
|- a : A]]$ regardless of whether $[[a]]$ is a term or a type. There
is no need to distinguish between big types and small types
and duplicate our typing specification.

@@

\section{Logical Relation}
\label{sec:logreldep}
\begin{figure}[h]
\drules[I]{$[[Interp I i A S]]$}{Logical Relation}{Void, Bool, Eq, Pi, Set, Red}
\caption{Logical relation for \lang}
\label{fig:logrel}
\end{figure}
In this section, we define the logical relation for \lang{} in the
form of an inductively defined relation (Figure~\ref{fig:logrel}).
Throughout the text, we use the notation $[[PowerSet S]]$ denotes the powerset of the set $[[S]]$.
The logical relation takes the form $[[Interp I i A S]]$. The
metavariables $[[A]]$ and $[[i]]$ stand for terms and natural
numbers respectively, as introduced earlier in
Figure~\ref{fig:syntax}.
The metavariables $[[I]]$ and $[[S]]$ are
sets with the following signatures.
\begin{equation*}
  \begin{split}
    [[I]] &\in [[ { j | j < i  } ->  PowerSet STm ]] \\
    [[S]] &\in [[PowerSet STm]]
  \end{split}
\end{equation*}
The function $[[I]]$ is a family of sets of terms indexed by
natural numbers strictly less than the parameter $[[i]]$, which
represents the current universe level.  In \rref{I-Set}, function
$[[I]]$ is used to define the meaning of
universes that are strictly smaller than the current level $[[i]]$. The
restriction $[[j < i]]$ in \rref{I-Set} is crucial for our system to
be predicative. Removing the ordering constraint would result in a
system where one can encode Girard's paradox~\citep{girard-thesis}.

\begin{figure}[h]
\begin{equation*}
    [[InterpR i A S]] := [[ Interp I i A S  ]], \text{where } [[I i]] := [[{A | exists S , InterpR i A S}]]
\end{equation*}
\caption{Logical relation for all universe levels}
\label{fig:logrelrec}
\end{figure}
To tie the knot and obtain an interpretation of all universe levels,
we define in Figure~\ref{fig:logrelrec} the final version of our interpretation judgment recursively
using the well-foundedness of the strict less than relation on natural
numbers (recall that
the parameter $[[I]]$ of $[[Interp I i A S]]$ takes only natural
numbers strictly less than $[[i]]$ as its input).
The judgment $[[InterpR i A S]]$ now reads that the type $[[A]]$ is a
level-$[[i]]$ type \emph{semantically} inhabited by terms from the set
$[[S]]$.

@@

To derive consistency, it suffices to restrict $[[S]]$ to the set of
terms to reduce to closed terms. \Rref{I-Void, I-Bool} are
unsurprising; when considering only closed terms, the empty type
should not be inhabited and therefore corresponds to the empty set,
whereas the boolean type is semantically inhabited by terms that
evaluate to the boolean values $[[true]]$ or $[[false]]$. \Rref{I-Eq}
says that an equality type $[[a ~ b : A]]$ corresonds to the
set of terms that evaluate to $[[refl]]$ when $[[a <=> b]]$ holds and
otherwise corresponds to the empty set. Side conditions like $[[a <=>
b]]$ are typically required for indexed types, of which equality types
are an instance.

>>>>

\newif\ifcomments     %% include author discussion
\newif\ifanonymous    %% include author identities
\newif\ifextended     %% include appendix
\newif\ifsubmission   %% prepare the submitted version
\newif\ifpublic       %% version available for posting / final version

\commentstrue         %% toggle comments here
\extendedfalse
\anonymoustrue

\submissionfalse     %% at most one of these must be true (neither for draft version)
\publicfalse          %% but if you want to see comments, these should both be off

% If we are going to make a version public, i.e. on arXiv, we should 
% make sure that there are no comments and our names are on it.
\ifpublic
\submissionfalse
\commentsfalse
\anonymousfalse
\fi

%% If we are submission, make sure there are no comments and our names 
%% are NOT on it.
\ifsubmission
\publicfalse
\commentsfalse
\anonymoustrue
\fi

%%
%% This is file `sample-acmsmall.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `acmsmall')
%%
%% IMPORTANT NOTICE:
%%
%% For the copyright see the source file.
%%
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-acmsmall.tex.
%%
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%%
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart} or whichever template is required
%% for your publication.
%%
%%
\documentclass[acmsmall,screen=true,
\ifpublic review=false\else,review=true\fi
  ,anonymous=\ifanonymous true\else false\fi]{acmart}
\usepackage{ottalt}
\usepackage{minted}
\usepackage{xspace}
\usepackage{tcolorbox}
\usepackage[para]{footmisc}
\definecolor{lightgray}{gray}{0.85}
\newcommand{\dotv}[2]{\href{#1}{\texttt{#1}}{\texttt{:#2}}}
\newcommand{\lang}{$\lambda^H$\xspace}
\inputott{rules}
%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

\usepackage{draft}

\ifcomments
\newnote{scw}{blue} % Stephanie Weirich
\newnote{yl}{purple} % Yiyun Liu
\else
\newcommand{\scw}[1]{}
\newcommand{\yl}[1]{}
\fi

@@

The consistency proofs of various dependently typed systems, including Martin-Lof's
type theory and the Calculus of Constructions, have long been
available in the literature. In particular, recent works such as \citet{nbeincoq},
\citet{decagda}, and \citet{martin-lof-a-la-coq} mechanize stronger
results\scw{stronger than what?} related to the decidability of type conversion or type
checking of dependently typed systems, from which consistency can be
derived as a corollary.

The underlying technique of the forementioned
works is proof by logical relation,
which involves interpreting types as reducibility predicates,
representing sets of terms satisfying certain properties with respect
to the reduction relation. While the proof technique and the
consistency result for dependent types are both well-established,
there is a lack of rigorous and accessible material that shows
how proof by logical relation can be applied to dependently typed
systems.
\scw{You are choosing to focus on mechanized logical relations proofs for 
dependent type theories, not consistency proofs for dependent type theories. 
Why? (And your title shuld match}

Introductory materials about logical relations such as
\citet{skorstengaard2019introduction}, \citet{harper2016practical}
start from logical relations for simply typed languages,\scw{proving 
results such as \ldots} and
eventually extend the technique to build a relational model for System
F in order to derive
parametricity.
\scw{Are there some OPLSS notes from Amal Ahmed to also cite? Other modern textbooks?
Does Aspinall and Hofmann's chapter of ATTAPL include a logical relation?}
\scw{Seems like you should also mention POPLmark reloaded somehere too?}
 \citet{harperkripke} gives a gentle introduction on how
logical relations for closed terms can be extended to talk about open
terms so we can derive properties such as normalization for well-typed
open terms. Overall, the introductory texts about logical relations
cover systems and properties with varying degrees of complexity, from
simply typed to polymorphicly typed, logical predicate to logical
equivalence, closed terms to open terms.

@@

The goal of this paper is to give a tutorial on proof by logical relation for
dependent types in a simple and digestable format. \scw{This sentence should come as 
early as possible.}
The system we present is small but covers
important features\scw{list these features explicitly: 
large eliminations, indexed types, others? \ldots}  that are only available or require special treatments in
dependently typed systems. We choose boolean
types over natural numbers as our base type for simplicity, but include an
intensional identity type in our type system to show how indexed
types are treated in a logical relational proof. \scw{Talk about 
identity type first.} Unlike
\citet{nbeincoq,decagda,martin-lof-a-la-coq} but similar to
\citet{anand2014towards}, we include an infinite hierarchy of
universes to not only support type-level computations, but also avoid
the unnecessary code duplication pointed out by \citet{nbeincoq} when
big and small types are treated non-uniformly.

@@

In Section~\ref{sec:extension}, \scw{what about sections 2-4?}

 we extend our logical
relation to prove the stronger property that every well-typed term
has a normal form. Most interestingly, we find that extending the
language with
$\eta$ rule for functions and the metatheoretic result to normalization for open terms are
mostly orthogonal to the fact that the language is dependently
typed. In our mechanized proof, the overall
structure of the logical relation remains unchanged, and the additional
proof obligations are mostly related to properties about the untyped
lambda calculus, all of which can be derived through syntactic
means.
In other words, once we have established the base technique for
proof by logical relation for dependent types, we can factor out the
complexities of an extension the are not specifically related to dependent
types; such an extension, if desired, may be studied in the context of
a simply typed language and later ported into a dependently typed setting.

@@

\scw{The end of this section needs to include an explicit summar of the 
contributions of the paper. }

@@

\mathit{Contexts}\\
[[G]]       & ::= & [[empty]]\ |\ [[G ++ A]] &  \\ \\
\mathit{Terms}\\
[[a]],[[b]],[[c]],[[t]],[[p]],[[A]],[[B]] & ::= & [[Set i]]\ |\ [[var i]]\  |\ [[Void]]
                  & \mbox{universes, variables, empty type} \\
            & |   & [[Pi A B]]\ |\ [[\ A a]]\ |\ [[a b]]
                  & \mbox{function types, abstractions, applications} \\
            & |   & [[a ~ b : A ]]\ |\  [[refl]]\ |\ [[J t a b p]]
                  & \mbox{equality types, reflexivity proof, J eliminator} \\
            & |   & [[Bool]]\ |\  [[true]]\ |\  [[false]]\ % |\  [[if a b0 b1]]
                  & \mbox{boolean type, true, false} \\
            & |   & [[if a b0 b1]]
                  & \mbox{if} \\ \\
\mathit{Renaming}\\
[[xi]] & \in & [[SNat -> SNat]] & \\ \\
\mathit{Substitution}\\
[[rho]] & \in & [[SNat -> STm]] &
\end{array}
\]
  \caption{Syntax of \lang \scw{Where does the name come from?}}
  \label{fig:syntax}
\end{figure}

@@

The syntax of \lang can be found in Figure~\ref{fig:syntax}. 
\scw{Why so many metavariables for terms? Do you need $c$, $p$ and $t$?}
We use
the unscoped de Bruijn representation for both our Coq development and
the informal presentation in our paper.  \scw{Give an overview of the figure}
\scw{Define a convention that separates your use of natural numbers. Sometimes they 
are universe levels and sometimes they are de Bruijn indices. Better to have separate
metavariables for each. (And it is possible to generalize universe levels to structures other than natural numbers. Ask Jonathan.}
We find de Bruijn
representation advantageous for specification since it leaves very
little ambiguity about the variable freeness side
conditions, making our proof more easily reproducible. Furthermore, as we discuss in
Section~\ref{sec:automation}, de Bruijn representation is much more
amenable to automated reasoning in proof assistants.
We adopt from \citet{autosubst} the notations for simultaneous
renaming, substitution, and other auxiliary definitions related to our
syntax, summarized in Figure~\ref{fig:auxdef}. 
\scw{Need to explain the notations in the text. What are these operations? 
What do readers need to understand about them? Help me understand the figure}
\scw{As this is a tutorial paper, you'll need to explain more about how de Bruijn
indices work.}
We omit most of the
definitions of renaming and substitution and only show the definition
of a few representative cases of substitution.
\begin{figure}[ht]
  \begin{equation*}
    \begin{split}
      \begin{array}{lll}
        \mathit{IdentityRen} \\
        [[id i]] & := & [[i]] \\ \\
        \mathit{IdentityTm} \\
        [[idtm i]] & := & [[var i]] \\ \\
        \mathit{ConsRen ([[xi .: n]])} \\
        ([[xi .: n]])([[0]]) & := & [[n]] \\
        ([[xi .: n]])([[Suc i]]) & := & [[xi i]] \\ \\
        \mathit{ConsSubst ([[rho .: a]])} \\
        ([[rho .: a]])([[0]]) & := & [[a]] \\
        ([[rho .: a]])([[Suc i]]) & := & [[rho i]] \\ \\
        \mathit{Curried\ Add} \\
        [[up n i]] & := & [[n]] + [[i]]
      \end{array}
    \end{split}
    \qquad \qquad
    \begin{split}
      \begin{array}{lll}
        \mathit{UpRen} \\
        [[( up xi ) 0]] & := & [[0]] \\
        [[( up xi ) Suc i]] & := & [[ Suc ren xi i ]] \\ \\
        \mathit{Renaming ([[a < xi >]])} \\
        [[(Pi A B) < xi >]] & := & [[Pi A < xi > B < up xi >]] \\
        [[(a b) < xi >]] & := & [[a < xi > (# b < xi > #)]] \\
        \ldots \\ \\
        \mathit{UpTm \scw{shifting? lifting?}} \\
        [[( up rho ) 0]] & := & [[var 0]] \\
        [[( up rho ) Suc i]] & := & [[ rho i < up 1 > ]] \\ \\
        \mathit{Substitution ([[ a { rho }  ]])} \\
        [[var i { rho }  ]] & := &  [[rho i]] \\
        [[(Pi A B) { rho }]] & := & [[Pi A { rho } B { up rho }]] \\
        [[(a b) { rho }]] & := & [[a { rho } (# b { rho } #)]] \\
        && \fbox{ \scw{case for $\lambda$ } } \\
        \ldots
      \end{array}
    \end{split}
  \end{equation*}
  \caption{Auxiliary Functions over Syntax}
  \label{fig:auxdef}
\end{figure}

@@

% Figure~\ref{fig:auxdef} shows the auxiliary definitions over the term
% syntax, including renaming, substitution, and operations over
% the typing context or substitution.
As a
dependent type theory, terms and types are collapsed into the same
syntactic category. Dependent functions take the form $[[Pi A B]]$ and
we use the notation $[[A -> B]]$ when the output type $[[B]]$ is not
dependent on the input variable. $[[Set i]]$ represents the universe
type where $[[i]]$ ranges over the set of natural numbers.
 Finally,
\lang\scw{Don't begin a sentence with a symbol} also includes an intensional identity type $[[a ~ b : A]]$ whose
proofs can be eliminated by the J-eliminator $[[J t a b p]]$, where
$[[p]]$ is an equality proof between $[[a]]$ and $[[b]]$, and $[[t]]$
is the term whose type is to be casted.
\scw{Example of what you can use identity types for? Or an example of 
a program that uses J?}

\lang is expressive enough to support large
eliminations, the ability to compute a type using a term as input. For
example, the function $[[\ Bool if var 0 Bool Bool -> Bool]]$ returns
either $[[Bool]]$ or $[[Bool -> Bool]]$ depending on whether the input
is $[[true]]$ or $[[false]]$.

@@

\scw{motivate parallel reduction. Why are you telling us this now. Is this 
an explanation of definitional equality and its properties?}
Figure~\ref{fig:par} shows the definition of the parallel reduction
relation, which takes the form $[[a => b]]$. We use $[[a =>+ b]]$ to
represent its transitive and reflexive closure, which in turn allows us to define
the coherence relation $[[a <=> b]]$.\scw{where is the terminology ``coherence''
from? I'm not familiar with it. Maybe you can replace it with ``definitional 
equivalence'' as that is how it is used in \lang?}
We say that two terms $[[a]]$
and $[[b]]$ are coherent if they can eventually reduce to some common
term $[[c]]$ through parallel reduction. The symmetric notation of
coherence suggests\scw{but you prove it below?} it is an equivalence relation.

@@

From Lemma~\ref{lemma:coherencerefl}, \ref{lemma:coherencesym}, and
\ref{lemma:coherencetrans}, we conclude that coherence is indeed an
equivalence relation. It is then trivial to derive that coherence is
exactly the same as untyped $\beta$-equivalence.\scw{cite barendregt?}

@@

Figure~\ref{fig:typing} gives the full typing rules for
\lang{}. The premises wrapped in \colorbox{lightgray}{gray} boxes can be shown to be
admissible syntactically, though some of them are required to
strengthen the inductive hypothesis of the fundamental theorem.
\scw{Is the precondition for T-ABS really admissible? Isn't it required 
to ensure predicativity? Have you proved it? Have you proved regularity?}
\scw{Do you have a formation rule for identity types?}

In \rref{T-Var}, $[[G i]]$ is the partial function defined
with the equations $[[ (G ++ A)  0 ]] := [[A]]$ and $[[ (G ++ A) Suc
i ]] := [[G i]] $.  \scw{explain the renaming in the conclusion of 
this lemma}
The precondition $[[i < | G | ]]$, where $[[ | G |
]]$ represents the length of the context $[[G]]$, ensures that $[[G
i]]$ has a defined output.

\Rref{T-Conv} uses the definition of coherence from earlier
as our equality judgment for type conversion. The use of an untyped
relation for type conversion makes our formulation slightly different
from languages such as MLTT~\citep{Martin-Lof-1973}, where the judgmental equality
takes the form $\Gamma \vdash a \equiv b : A$, which implies $[[G |- a :
A]]$ and $[[G |- b : A]]$, the well-typedness of $[[a]]$ and $[[b]]$.
\citet{siles2012pure} shows the equivalence of Barendregt's
Pure Type System~\citep{barendregt1991introduction}, which employs
untyped equality, and its variant that uses typed judgmental
equality. It is trivial to embed a system with judgmental
equality to a system with untyped equality by erasing typing
information. As a result, it is easy to port our consistency result from our
type system to a variant with typed judgmental equality
as long as the typed system does not include $\eta$ laws
that would require type annotations (e.g. the $\eta$-law for unit
types). % and we discuss in Section~\ref{sec:extension} how we can extend
% our proof to handle those rules.
\scw{Maybe move more of this paragraph and the next to the discussion/related
  work section, and add a forward pointer here.}
\scw{Just say that like Coq, the definition of the system uses untyped 
equality}

@@

Finally, since our system has an infinite universe hierarchy, we can
present the system à la Russell by using the same judgment form $[[G
|- a : A]]$ regardless of whether $[[a]]$ is a term or a type. There
is no need to distinguish between big types and small types
and duplicate our typing specification.
\scw{Is there an advantage for Tarski universes even with infinite 
hierarchy? or no?}

@@

\section{Logical Relation}
\scw{This section pre-supposes that we want to define a logical relation, 
but doesn't precisely state what we want to use it to prove, and why a logical
relation is suitable. (And why the property you want to prove is difficult to 
show!) Should add more motivation here.}

\label{sec:logreldep}
\begin{figure}[h]
\drules[I]{$[[Interp I i A S]]$}{Logical Relation}{Void, Bool, Eq, Pi, Set, Red}
\caption{Logical relation for \lang}
\label{fig:logrel}
\end{figure}
In this section, we define the logical relation for \lang{} in the
form of an inductively defined relation (Figure~\ref{fig:logrel}).
Throughout the text, we use the notation $[[PowerSet S]]$ denotes the powerset of the set $[[S]]$.
The logical relation takes the form $[[Interp I i A S]]$. The
metavariables $[[A]]$ and $[[i]]$ stand for terms and natural
numbers respectively, as introduced earlier in
Figure~\ref{fig:syntax}.
\scw{Because the type of $I$ depends on $i$, I'd prefer a syntax that 
keeps them in order. }
\scw{Many introductory texts define the relation as a recursive function 
over type structure, or step-indices. You use an inductive relation instead, why?}
\scw{Can we view this inductive relation as the graph of the partial function
that is defined recursively over types?}
\scw{What is this form extensible too? impredicative quantification? recursive 
types? }
\scw{Is it worth observing here that this definition is not over sets of typed
terms. That it characterizes all terms that look like booleans (i.e. evaluate to 
true or false) or all terms that look like proofs (i.e. evaluate to refl). The 
fact that there is no connection between p and a and b in the I-Eq case is strange
looking. Need to explain.  }

The metavariables $[[I]]$ and $[[S]]$ are
sets with the following signatures.
\begin{equation*}
  \begin{split}
    [[I]] &\in [[ { j | j < i  } ->  PowerSet STm ]] \\
    [[S]] &\in [[PowerSet STm]]
  \end{split}
\end{equation*}
The function $[[I]]$ is a family of sets of terms indexed by
natural numbers strictly less than the parameter $[[i]]$, which
represents the current universe level.  In \rref{I-Set}, function
$[[I]]$ is used to define the meaning of
universes that are strictly smaller than the current level $[[i]]$. The
restriction $[[j < i]]$ in \rref{I-Set} is crucial for our system to
be predicative. Removing the ordering constraint would result in a
system where one can encode Girard's paradox~\citep{girard-thesis}.
\scw{Spell this out a bit more about what goes wrong. Where does the 
proof break? Also spell out how this enforces predicativity.}

\begin{figure}[h]
\begin{equation*}
    [[InterpR i A S]] := [[ Interp I i A S  ]], \text{where } [[I i]] := [[{A | exists S , InterpR i A S}]]
\end{equation*}
\caption{Logical relation for all universe levels}
\label{fig:logrelrec}
\scw{I didn't understand this figure without reading the text below.}
\end{figure}
To tie the knot and obtain an interpretation of all universe levels,
we define in Figure~\ref{fig:logrelrec} the final version of our interpretation judgment recursively
using the well-foundedness of the strict less than relation on natural
numbers (recall that
the parameter $[[I]]$ of $[[Interp I i A S]]$ takes only natural
numbers strictly less than $[[i]]$ as its input).
The judgment $[[InterpR i A S]]$ now reads that the type $[[A]]$ is a
level-$[[i]]$ type \emph{semantically} inhabited by terms from the set
$[[S]]$.

@@

To derive consistency, it suffices to restrict $[[S]]$ to the set of
terms that reduce to closed terms. \Rref{I-Void, I-Bool} are
unsurprising; when considering only closed terms, the empty type
should not be inhabited and therefore corresponds to the empty set,
whereas the boolean type is semantically inhabited by terms that
evaluate to the boolean values $[[true]]$ or $[[false]]$. \Rref{I-Eq}
says that an equality type $[[a ~ b : A]]$ corresonds to the
set of terms that evaluate to $[[refl]]$ when $[[a <=> b]]$ holds and
otherwise corresponds to the empty set. Side conditions like $[[a <=>
b]]$ are typically required for indexed types, of which equality types
are an instance.
